Automatic discovery of cell types and microcircuitry from neural connectomics

### 2014-06-04 docs/paper.tex
Moving experiments to here to generate final versions for paper

%\documentclass{nature}
\documentclass{article}
\usepackage[english,american]{babel}
\usepackage[backend=biber,style=nature]{biblatex}
\usepackage[disable]{todonotes}
\usepackage{grffile}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{graphicx}
\usepackage{fullpage}
%\usepackage{algorithm}
%\usepackage{algorithmic}
\usepackage{amsmath}
%\usepackage{subcaption}
%\usepackage{multicol}
\usepackage{csquotes}
\usepackage{pdfpages}
\usepackage{placeins}

@@

\title{Automatic discovery of circuit structure from neural connectomics}
\author{Eric Jonas$^1$, Konrad Kording$^{2, 3,4}$}

@@

\begin{abstract}
  New techniques produce massive data about neural connectivity,
  necessitating new analysis methods to discover underlying
  structure. It has long been assumed that discovering the local
  patterns of microcircuitry is the key to understanding neural
  function. Here we developed a nonparametric Bayesian technique that
  identifies neuron types and circuitry patterns in connectomics
  data. We show that the approach recovers known neuron types in the
  retina, reveals interesting structure in the nervous system of
  c. elegans, and automatically discovers the structure of
  microprocessors. Our approach extracts structural meaning from
  connectomics, enabling new approaches of deriving anatomical
  insights from these emerging datasets.

\end{abstract}

@@

The brain consists of multiple kinds of neurons, each of which is
hypothesized to have a specific role in the overall
computation. Neuron types differ in many ways, e.g. chemical or
morphological, but they also differ in the way they connect to one
another. In fact, the idea of well defined local connectivity patterns
has been prominent in many areas, ranging from sensory (e.g. retina,
\autocite{Masland2001} via processing (e.g. neocortex \autocite{Mountcastle1997})
to movement (e.g. spinal cord) \autocite{Grillner2005}. Indeed, such
repeated patterns also exists in human-made computing circuits. It
remains an important challenge to develop algorithms to use anatomical
data, e.g. connectomics, to back out underlying connectivity motifs.

@@

The discovery of structure in probabilistic graphs is a crucial aspect
in machine learning. Commonly used algorithms include
community-based-detection methods \autocite{Girvan2002}, and
stochastic block models \autocite{Nowicki2001}.  While these
approaches can incorporate the probabilistic nature of neural
connections \autocite{Hill2012} they do not generally consider the
known fact that the probability of connections between neurons depends
on distance. Combining such probabilistic models with a spatial
component to model distance promises to improve the detection of
microcircuitry from neural data.

When it comes to neuroscience and other computing systems, we expect
patterns of connectivity much more complex than traditional motifs,
exhibiting a strong spatial dependence arising from the complex
genetic, chemical, and activity-based neural development process.

Here we describe a Bayesian non-parametric model that can discover
circuit structure automatically from connectomics data: both the cell
types and their spatial patterns of interconnection. We apply it to
three computing systems: the mouse retina connectome
\autocite{Helmstaedter2013}, the c. elegans connectome
\autocite{White1986}, and a ``connectome'' of a classical
microprocessor \autocite{James2010}. Note we do not explicitly
consider cellular morphology, absolute position (cell layer), or any
chemical or genetic information. Yet in all cases, we discover
cell-type to cell-type connectivity that depends on spatial distance
and these connectivity functions depend on the two involved cell
types. Comparing the cell types discovered by the algorithms with
those obtained by anatomists based on orthogonal information reveals a
high degree of agreement. We present a scalable approach to infer
microcircuitry from connectivity data.

\begin{figure}
  \centering 
  \centerline{\includegraphics[width=183mm]{overview.ai}}
  \caption{An overview of our method. a. Input data is
a connectivity matrix indicating whether cell $i$ (row)
synapses on cell $j$ (column). The location of each cell in space is known. 
b. Our algorithm discovers hidden cell types in this connectivity data
by assuming all cells of a type share a distance-dependent connectivity 
profile, c.), with cells of other types. This results in a clustering 
of the cells by those hidden types. b.) shows the cell connectivity
matrix with cells of the same type grouped together. d.) shows
the learned probability of connection between our different types
at various distances, illustrating short-range and long-range
connectivity patterns.}
\label{fig:overview}
\end{figure}

\section{Results}
We use a Bayesian structure discovery approach that is based on
prior knowledge about the way positions and cell types affect the
probability of connections between cells, a so-called generative
model. We term our model the ``infinite spatial-relational model'' (iSRM), 
in keeping with the existing nonparametric literature.  
The probability of connections between pairs of neurons
decreases with distance, but how high the baseline probability of
connection is and how quickly it decays with distance depends on the
two cell-types. Note that this simplistic prior information is all that 
we tell our model; no additional physiological information is included.

We combine annealing with Markov Chain Monte Carlo to
perform simultaneously joint posterior inference over the structure,
parameters, and hyperparameters of the model (see Methods). This
procedure automatically learns the number of cell types, which cells 
belong to which types, along with the 
the distance and type dependent connectivity probabilities (see
fig~\ref{fig:overview}). These results can then be compared with the
results of previous anatomical studies that have used other aspects of
neurons, such as chemical properties and cell-morphology to divide
neurons into types.

Before we can use a probabilistic model to analyze real data we first
need to assess if the model works properly. We thus started out by
simulating data for which we know the correct structure and comparing
the estimated structure based on the algorithm (see methods) with the
one we used for simulation. We find that the model does a good job of
recovering the structure we used to generate the data
(fig~\ref{fig:synthetic}a) Indeed, it does far better than a simpler
and well established model (the infinite stochastic block model, or
iSBM) that assumes that there is no distance dependence to the
connectivity. The model converges relatively quickly to an estimate of
the most probable values for the cell types, which is enabled by using
a combination of simulated annealing and parallelized Markov chain
Monte Carlo (see methods for details). It appears that the model
behaves the way it should and allows application to relatively large
datasets.

\begin{figure}
  \centering 
  \centerline{\includegraphics{synthetic.ai}}
  \caption{Correct recovery of true numbers of hidden classes
in synthetic data. a.) Other models over-cluster the number
of classes as they fail to take distance into account, whereas
our model finds close to the true number of classes. b.) ARI, a measure
of ``correct clustering'', for
different true class counts, between our model and a traditional
iSBM. c.) For synthetic data, 
the iSBM overclusters, finding a large number of spatially-localized types. 
In contrast, the infinite spatial-relational model recovers the true spatial
extent of the underlying types.}
\label{fig:synthetic}
\end{figure}

@@

\begin{figure}
  \centering 
  \centerline{\includegraphics{mouseretina.ai}}
  \caption{Discovering cell classes in the mouse retina connectome. 
a.) Input connectivity data for 950 cells for which soma positions were known. b. clustered connectivity matrix. c.) connectivity diagram showing our clusters, as
well as the cell depth and known cell type of each cell. Our model agrees
with the cell types traditionally identified by anatomists, and identifies 
laminar-specific connectivity patterns. d.) Connectivity between our
clusters as a function of distance -- the cluster consisting primarily of
retinal ganglion cells (lower left-center) exhibits the expected near and
far connectivity. e. The spatial distribution of our cell types -- each cell type tesselates space.}
\label{fig:mouseretina}
\end{figure}

The retina of the mouse \autocite{Masland2001} is an example where we
should expect to have connectivity patterns that are well approximated
by our generative model. It is known that there are multiple classes
of cells that can be grouped into ganglion cells that transmit
information to the rest of the brain, bipolar cells that connect
between different cells, and amacrine cells that feed into the
ganglion cells \autocite{cite some review}. Recent research
\autocite{Helmstaedter2013} has produced a large dataset containing
both the types of cells from orthogonal approaches, and also the
connectivity matrix between all reconstructed cells. We want to know
to which level we can classify neurons into types based on this
connectivity only.

The algorithm took 8 hours to converge to a locally optimal solution,
dividing neurons into a set of cell types (fig~\ref{fig:mouseretina}c, each
wedge is a type). For each pair of neurons there is a specific
distance dependent connection probability (fig~\ref{fig:mouseretina}b,c,d),
which is well approximated by the model fit. Moreover, each type of
cell is rather isotropically distributed across space
(fig~\ref{fig:mouseretina}e) as should be expected for true cell types. The
algorithm is thus able to extract meaningful structure from the data.

@@

\todo{Do we assign any of the unknown cells correctly?  are types 72-79
“different” or simply unknown?}

We compared the results of our model to the traditional iSBM to see if
the addition of spatial information was necessary. We find that, as
expected from our synthetic experiments, the iSBM massively
over-segmented the data, finding spatially-localized “communities” of
cells which bore little relation to known types.  The average spatial
extent of the types found by the iSBM was $14.6\pm8.1 \mu m$. However,
in the ground truth by anatomists the average distance was $34.3\pm7.6
\mu m$ and $35.3\pm 2.4 \mu m$, quite close to the $34.5\pm 1.0 \mu m$
discovered by our algorithm. Modeling the distance-dependence prevents
over-segmentation and thus allows more meaningful estimates.

@@

The roundworm Caenorhabditis elegans is a model system in
developmental neuroscience\autocite{White1986}, with the location and
connectivity of each of 302 neurons deterministically known. This
incredible cross-organism consistency led to early mapping of the
c. elegans chemical and electrical (gap junction) connectome. Unlike
the retina, only the motor neurons in c. elegans exhibit regular
distribution in space, being distributed along the body axis. Most
interneurons are concentrated in various ganglia that project
throughout the entire animal, and the sensory neurons are primarily
located in a small number of anterior ganglia.

\begin{figure}
  \centering 
  \centerline{\includegraphics[width=5in]{celegans.ai}}
  \caption{Discovering cell types in the \textit{c. elegans}
    connectome. a.) Initial cell connectivity matrix, red are
    (directional) chemical synapses, blue are electrical
    synapses. Point size is proportional to synapse count between
    those cells. b.) Clustered adjacency matrix showing identified
    clusters. c.) Inter-cluster connectivity, showing the position of
    each cell along the body axis as well as the ``true'' cell type --
    both coarse (motor, sensory, and interneuron) as well as fine
    (small text labels). }
  \label{fig:celegans}
\end{figure}

@@

We identify cell types that reflect the known motor/non-motor neuron
classes, even though this system lacks the strong repeat
microcircuitry our model was designed for.  Motor-neuron types AS, DA,
and VA, all exclusively postsynaptic, are clustered together, as are
motor-neuron types VD and DD. Traditional types VC, DB, and VB also
mostly share a cluster. Various head motor neurons, including types
SMD, RMD, and RIM are clustered together. Interneurons with known
anatomically-distinct connectivity patterns, such as RIA (2 cells) or
AVA (2 cells) are clustered into pure types. The algorithm even
correctly places the single-cell type DVB by itself. Note our
clustering does not perfectly reflect known divisions -- several
combinations of head and sensory neurons are combined, and a
difficult-to-explain group of 2 VD, 2 VB, a VA, and a DD neuron are
grouped together. Our identified cell types thus reflect a
``coarsening'' of known types, based entirely on connectivity and
distance information, even when the organism exhibits substantially
less spatial regularity than the retina.

\subsection{Microprocessor}
\begin{figure}
  \centering 
  \centerline{\includegraphics{mos6502.ai}}
  \caption{Discovering transistor types in the MOS 6502
    microprocessor. a.) The clusters and connectivity versus distance
    for connections between Gate and C1, Gate and C2, and C1 and C2
    terminals on a transistor. Red and teal types have a terminal
    pulled down to ground and mostly function as inverters. Purple
    class are clocked, stateful transistors, orange control the ALU
    and yellow control the special data bus. b.) The spatial
    distribution of the transistors in each cluster. }
\label{fig:mos6502}
\end{figure}

@@

\section{Discussion}
We have presented a machine learning technique that allows cell types
and microcircuitry to be discovered using only distance and
connectivity data. We have shown its applicability to a newly
published connectomics dataset from the mouse retina, the classical
connectivity of the worm c. elegans, and the connectivity of a
historical man-made microprocessor. We have found that distance and
connectivity alone are sufficient to discover cell types and
microcircuitry that were known to exist in the systems based on decades
of previous research which had utilized extensive additional
information.

For probabilistic models like ours, no known solution exists to
exactly find the most probable parsing of the neurons into cell-types
and connectivity patterns . We employ a collection of Markov-chain
Monte carlo techniques (see Methods) but while different
initializations converge to similar ultimate values, we can never
realistically obtain the global optimum. There are a broad range of
techniques that may offer better approximations to the global optimum
\autocite{Wanga2012} and future work could adapt them to
the problem we addressed here.

Our inference becomes slower as the amount of data increases. Our
algorithm required 30 hours for 1000 neurons and could take months for
substantially longer datasets. Scaling this class of probabilistic
model is an active area of research, and recent results in both
variational methods \autocite{Hoffman2013} and spectral learning
  \autocite{Anandkumar2012} promise substantial speed increases.

  Larger datasets will allow algorithms to distinguish more distinct
  types \todo{Add supplemental data} and we expect closer agreement
  with existing anotomical knowledge as more data become available.
  Moreover, in general, for such problems precision increases with the
  size of the dataset and the cells that we have are not sufficient to
  statistically distinguish all the cell types known in anatomy (such
  as the $\sim 70$ in the retina). Still, using only connectivity and
  distance it is possible to meaningfully divide neurons into types.

@@

There exist a range of previous approaches to the discovery of
neural microcircuitry \todo{Do we want a citation here?}. These
generally involve a great deal of manual labor and ad-hoc
determination of what constitutes a “type” of cell -- to this day
there are disagreements in the literature as to the “true” types in
the mammalian retina. Much as phylogenomics has changed our
understanding of animal ontologies, modern large scale data will allow
the efficient unbiased discovery of cell types and circuits. The sheer
amount of available data demands the introduction of algorithmic
approaches.

The development of automatic identification and quantification of cell
type may also provide a new computational phenotype for quantifying
the effect of disease, genetic interventions, and
developmentally-experienced neural activity. Our method can in
principle identify neuron-types across non-connected graphs,
e.g. across animals. For example, the types of neurons in one animal
can be associated with the types of neurons in another animal, in the
same way as this is already possible through molecular markers
\autocite{Brown2009}. This could be particularly important if cell
types appear that are due to properties of the stimuli and experience
as opposed to just the molecular properties of cells, such as color
and orientation selective types in primary visual cortex
\autocite{Sincich2005,Lennie2005}. This would allow comparative
quantitative anatomy across animals, and aid the search for the
ultimate causes of connectivity.

Ultimately, we will need to combine information across different
sources into a joint understanding of neural function. Distinct cell
types differ in morphology, connectivity, transcriptomics, relation to
behavior or stimuli and many other ways. Algorithms based on each type
information can then be combined to synthesize all the available
information from one experiment or even across experiments into a
joint model of brain function.

@@

@@

@@

\newpage
\section{Methods }

@@

@@

@@

To pick the MAP, we run up to $100$ simultaneous chains in parallel,
initialized from different random initial points in the state space,
and then pick the chain with the highest log likelihood. Early
experiments attempted to use the full ensemble of chains as an
estimate of the posterior but were abandoned due to the difficulty of
summarizing the posterior over such a complex state space, which is an
ongoing area of research in the probabilistic modeling community
\todo{cite more? missing}.

\todo{Validation: do we want to write extensively about posterior
  enumeration tests, etc. ? }

@@

% \begin{addendum}
%  \item Put acknowledgements here.
%  \item[Competing Interests] The authors declare that they have no
% competing financial interests.
%  \item[Correspondence] Correspondence and requests for materials
% should be addressed to A.B.C.~(email: myaddress@nowhere.edu).
% \end{addendum}
\newpage
\section{Supplemental}

>>>>

\documentclass{nature}
\documentclass{article}
\usepackage[english,american]{babel}
\usepackage[backend=biber,style=nature]{biblatex}
\usepackage[disable]{todonotes}
\usepackage{grffile}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{graphicx}
\usepackage{fullpage}
%\usepackage{algorithm}
%\usepackage{algorithmic}
\usepackage{amsmath}
%\usepackage{subcaption}
%\usepackage{multicol}
\usepackage{csquotes}
\usepackage{pdfpages}
\usepackage{placeins}

@@

\todo{“Anatomist cell type” text}
\todo{Consistent line thickness}
\todo{on adjacency matrix remove text}

\title{Automatic discovery of cell types and microcircuitry from neural connectomics}
\author{Eric Jonas$^1$, Konrad Kording$^{2, 3,4}$}

@@

\begin{abstract}
  Connectomics produces massive amounts of data,
  necessitating new analysis methods to discover the biological and
  computational structure. It has long been assumed
  that discovering neuron types and their relation to microcircuitry is crucial to
  understanding neural function. Here we developed a nonparametric
  Bayesian technique that identifies neuron types and microcircuitry
  patterns in connectomics data. It combines the information traditionally used by biologists, including connectivity, cell body location and the spatial distribution of synapses.  We show that the approach recovers known neuron types in the retina and enables predictions of connectivity, better than simpler algorithms. It also can reveal interesting structure in
  the nervous system of c. elegans, and automatically discovers the
  structure of a microprocessor. 
 Our approach extracts structural meaning from connectomics, enabling new approaches of automatically deriving  anatomical insights from these emerging datasets.
\end{abstract}

@@

The brain consists of multiple kinds of neurons, each of which is
hypothesized to have a specific role in the overall
computation. Neuron types differ in many ways, e.g. chemical or
morphological, but they also differ in the way they connect to one
another. In fact, the idea of well defined local connectivity patterns
(microcircuits) has been prominent in many areas, ranging from sensory
(e.g. retina, \autocite{Masland2001} via processing (e.g. neocortex
\autocite{Mountcastle1997}) to movement (e.g. spinal cord)
\autocite{Grillner2005}. Indeed, such repeated patterns also exists in
human-made computing circuits. It remains an important challenge to
develop algorithms to use anatomical data, e.g. connectomics, to automatically back
out underlying microcircuitry.

@@

The discovery of structure in probabilistic graphs is a crucial aspect
in machine learning. Commonly used algorithms include
community-based-detection methods \autocite{Girvan2002}, and
stochastic block models \autocite{Nowicki2001}.  While these
approaches can incorporate the probabilistic nature of neural
connections \autocite{Hill2012} they do not incorporate the additional
richer structure present in connectomics data -- the location of cell
bodies, the spatial distribution of synapses, and the distances
between neurons. Of particular importance is that the probability of
connections has a strong spatial component, a factor that is hard to
reconcile with many other methods. A model attempting to fully capture
the variation in the nervous system should take into account the broad
set of available features.

When it comes to neuroscience and other computing systems, we expect
patterns of connectivity much more complex than traditional motifs,
exhibiting a strong spatial dependence arising from the complex
genetic, chemical, and activity-based neural development processes.

Here we describe a Bayesian non-parametric model that can discover
circuit structure automatically from connectomics data: the cell
types, their spatial patterns of interconnection, and the locations of
somata and synapses. We primarily focus on the recently-released mouse
retina connectome \autocite{Helmstaedter2013}, but additionally
examine the c. elegans connectome \autocite{White1986}, and then
``connectome'' of a classical microprocessor
\autocite{James2010}. Comparing the cell types discovered by the
algorithms with those obtained manually by human anatomists reveals a
high degree of agreement. We present a scalable approach to infer
microcircuitry from connectomics data.

\begin{figure}
  \centering 
  \centerline{\includegraphics[width=183mm]{fig1b.ai}}
  \caption{An overview of our method. a. Input data is a connectivity
    matrix indicating whether cell $i$ (row) synapses on cell $j$
    (column), a distance matrix between cell soma, b. the depth of the
    cell soma, and the c. depth profile of the cells synaptic
    connectivity. d. Our algorithm discovers hidden cell types in this
    connectivity data by assuming all cells of a type share a
    distance-dependent connectivity profile, similar depth, and a
    similar synpatic density profile, with cells of other types.  This
    results in a clustering of the cells by those hidden
    types. d. shows the cell connectivity matrix with cells of the
    same type grouped together. e.) shows the learned probability of
    connection between our different types at various distances -- in
    this case, the cells are likely to connect when they are
    close. f. shows a connectivity profile with virtually no
    probability of connectivity. We can plot this these types g.) to
    show the relationship and spatial connections, and how probable
    various types are to connect to one another. h. The connectivity
    between our discovered types changes as a function of distance. }

\label{fig:overview}
\end{figure}

\section{Results}
We start with the generic assumption that cell types exist and affect
soma depth, distribution of synapses, as well as a cell type and
distance dependent connection probability. From this assumption we
develop a Bayesian algorithm that estimates the underlying cell types.
We start with a model for connectivity, the infinite stochastic block
model (iSBM)\autocite{Kemp2006a,Xu2006}, which has been shown to
meaningfully cluster connection graphs. We extend this approach by
adding distance dependence to model salient aspects of
microcircuitry. We additionally model cell body depth as a mixture of
Gaussians and each cell’s synapse density profile as mixture of
mixtures of Gaussians. For example, retinal ganglion cells may synapse
on nearby, but not far away, amacrine cells, with ganglion cells being
superficial and having a broad distribution of synapses.  This
procedure automatically learns the number of cell types, which cells
belong to which types, along with the the distance and type dependent
connectivity probabilities (see fig~\ref{fig:overview}). We apply our
algorithm to datasets from mouse retina, c-elegans and a historical
microprocessor. Anatomists classify cells based on many features,
giving us a meaningful baseline to compare against.

Before we can use a probabilistic model to analyze real data we first
need to assess if the model works properly. We thus started out by
simulating data for which we know the correct structure and comparing
the estimated structure based on the algorithm (see methods) with the
one we used for simulation. We find that the model does a good job of
recovering the correct number of cell types,
(fig~\ref{fig:synthetic}a), the the cell identities
(fig~\ref{fig:synthetic}b), and the spatial extent of each type
(fig~\ref{fig:synthetic}c).  Indeed, it does far better than the
existing infinite stochastic block model that assumes that there is no
distance dependence to the connectivity. The model converges
relatively quickly to an estimate of the most probable values for the
cell types, which is enabled by using a combination of simulated
annealing and parallelized Markov chain Monte Carlo (see methods for
details). It appears that the model behaves the way it should and
allows application to relatively large datasets.
<XXX now add extra features>

\begin{figure}
  \centering 
  \centerline{\includegraphics{synthetic.ai}}
  \caption{Correct recovery of true numbers of hidden types in
    synthetic data when incorporating spatial information. a.) The
    iSBM (which only uses connectivity information) over-estimates the
    number of classes as it fails to take distance into account,
    whereas our modeling of the combination of distance and
    connectivity finds close to the true number of classes. b.) ARI, a
    measure of ``correct clustering'', for different true class
    counts, between our model and a traditional iSBM. c.) For
    synthetic data, type discovery based only on connectivity are
    tiny spatially-localized groups -- in
    contrast, our model recovers the true spatial extent of the
    underlying types.  }
\label{fig:synthetic}
\end{figure}

@@

\begin{figure}
  \centering 
  \centerline{\includegraphics{mouseretina.ai}}
  \caption{Discovering cell classes in the mouse retina connectome.
    a.) Input connectivity data for 950 cells for which soma positions
    were known. b. clustered connectivity matrix. c.) connectivity
    diagram showing our clusters, as well as the cell depth and known
    cell type of each cell. Our model shows moderate agreement with
    the cell types traditionally identified by anatomist
    laminar-specific connectivity patterns. \todo{Add more text here}
    d.) Connectivity between our clusters as a function of distance --
    the cluster consisting primarily of retinal ganglion cells (lower
    left-center) exhibits the expected near and far connectivity. e.)
    The spatial distribution of our cell types -- each cell type
    tesselates space.}
\label{fig:mouseretina}
\end{figure}

The retina of the mouse \autocite{Masland2001} is an example where we
should expect to have connectivity patterns that are well approximated
by our generative model. It is known that there are multiple classes
of cells that can be grouped into ganglion cells that transmit
information to the rest of the brain, bipolar cells that connect
between different cells, and amacrine cells that feed into the
ganglion cells. Recent research
\autocite{Helmstaedter2013} has produced a large dataset containing
both the types of cells from orthogonal approaches, and also the
connectivity matrix between all reconstructed cells. We want to know
to which level we can classify neurons into types based on this
connectivity only.

The algorithm took less than 2 hours to converge to a good solution,
dividing neurons into a set of cell types (fig~\ref{fig:mouseretina}c, each
wedge is a type). For each pair of neurons there is a specific
distance dependent connection probability (fig~\ref{fig:mouseretina}b,c,d),
which is well approximated by the model fit. Moreover, each type of
cell is rather isotropically distributed across space
(fig~\ref{fig:mouseretina}e) as should be expected for true cell types. The
algorithm is thus able to extract meaningful structure from the data.

@@

<xxx discuss the various algorithms and plus and minus 
   Other model classes
   brielfly explain link prediction accuracy and ARI
   show how it compares to human example	
   the more humanlike the information it gets the more humanlike its classification but many techniques allow good link prediction
>

We compared the results of our model to the traditional iSBM to see if
the addition of spatial information was necessary. We find that, as
expected from our synthetic experiments, the iSBM massively
over-segmented the data, finding spatially-localized “communities” of
cells which bore little relation to known types.  The average spatial
extent of the types found by the iSBM was $14.6\pm8.1 \mu m$. Using
the anatomists' ground truth, the mean spatial extent of types was
$34.3\pm7.6 \mu m$, quite close to the $34.5\pm 1.0 \mu m$ discovered
by our algorithm. Modeling the distance-dependence prevents
over-segmentation and thus allows more meaningful estimates.

@@

The roundworm Caenorhabditis elegans is a model system in
developmental neuroscience\autocite{White1986}, with the location and
connectivity of each of 302 neurons developmentally determined,
leading to early measurement of the connectome. Unlike the retina,
only the motor neurons in c. elegans exhibit regular distribution in
space -- along the body axis. Most interneurons are
concentrated in various ganglia that project throughout the entire
animal, and the sensory neurons are primarily located in a small
number of anterior ganglia.

\begin{figure}
  \centering 
  \centerline{\includegraphics[width=7in]{otherdata.ai}}
  \caption{Discovering connectivity and type in both
    \textit{c. elegans} and the MOS 6502 microprocessor.  a.) Initial
    \textit{c. elegans} cell connectivity matrix, red are
    (directional) chemical synapses, blue are electrical
    synapses. Point size is proportional to synapse count between
    those cells. b.) Clustered adjacency matrix showing identified
    clusters. c.) Inter-cluster connectivity, showing the position of
    each cell along the body axis as well as the ``true'' cell type --
    both coarse (motor, sensory, and interneuron) as well as fine
    (small text labels). a.) The clusters and connectivity versus
    distance for connections between Gate and C1, Gate and C2, and C1
    and C2 terminals on a transistor. Red and teal types have a
    terminal pulled down to ground and mostly function as
    inverters. Purple class are clocked, stateful transistors, orange
    control the ALU and yellow control the special data bus. b.) The
    spatial distribution of the transistors in each cluster.}
  \label{fig:otherdata}
\end{figure}

@@

We identify cell types that reflect the known motor/non-motor neuron
classes, even though this system lacks the strong repeat
microcircuitry our model was designed for.  Motor-neuron types AS, DA,
and VA, all exclusively postsynaptic, are identified as a common type,
as are motor-neuron types VD and DD. Traditional types VC, DB, and VB
also mostly share a cluster. Various head motor neurons, including
types SMD and RMD, are clustered together. Interneurons with known
anatomically-distinct connectivity patterns, such as AVA (2 cells),
are clustered into pure types.The algorithm even correctly places the
single-cell types DVB and DVC by themselves.

Note our clustering does not perfectly reflect known divisions --
several combinations of head and sensory neurons are combined, and a
difficult-to-explain group of mostly VB and DB motor neuron types,
with VC split between various groups. Our identified cell types thus
reflect a ``coarsening'' of known types, based entirely on
connectivity and distance information, even when the organism exhibits
substantially less spatial regularity than the retina.

\subsection{Microprocessor}

@@

TODO FIGURE 5: 
We know we want the bar graph with ARI and AUC at the same time:
Conn only
conn and dist
conn and dist and synapse profile and depth
LPCM 
Just cell depth 
human dashed line (for link prediction)

\section{Discussion}
We have presented a machine learning technique that allows cell types
and microcircuitry to be discovered from connectomics data.  We have
shown its applicability to a newly published connectomics dataset from
the mouse retina. When compared to existing methods, we show how the
incorporation of all of this data yields results that combine both
high link-prediction accuracy and high agreement with human
anatomists.  I also reveals interesting structure when applied to the
classical connectivity of the worm c. elegans, and the connectivity of
a historical man-made microprocessor. We have found that combining the
available data types allows us to discover cell types and
microcircuitry that were known to exist in the systems based on
decades of previous research and allows good prediction of
connectivity.

For our probabilistic models, no known solution exists to
exactly find the most probable parsing of the neurons into cell-types
and connectivity patterns. We employ a collection of Markov-chain
Monte carlo techniques (see Methods) but while different
initializations converge to similar ultimate values, we can never
realistically obtain the global optimum. There are a broad range of
techniques that may offer good approximations to the global optimum
and future work could adapt them to
find more precise solutions to our problem.

For our probabilistic model, inference becomes slower as the amount of
data increases. Our algorithm required several hours for 1000
neurons. Scaling this class of probabilistic model is an active area
of research, and recent results in both variational methods
\autocite{Hoffman2013} and spectral learning \autocite{Anandkumar2012}
and future work could adapte them to find faster approximate solutions
to our problem.

Larger datasets will allow algorithms to distinguish more distinct
types and we expect closer agreement with existing anatomical
knowledge as more data become available.  Moreover, in general, for
such problems precision increases with the size of the dataset and the
cells that we have are not sufficient to statistically distinguish all
the cell types known in anatomy (such as the $\sim 70$ in the
retina). Still, using only connectivity and distance it is possible to
meaningfully divide neurons into types.

@@

There exist a range of previous approaches to the discovery of
neural microcircuitry \todo{ xxx as many cites as possible}. These
generally involve a great deal of manual labor and ad-hoc
determination of what constitutes a “type” of cell -- to this day
there are disagreements in the literature as to the “true” types in
the mammalian retina. Much as phylogenomics has changed our
understanding of animal ontologies, modern large scale data will allow
the efficient unbiased discovery of cell types and circuits. The sheer
amount of available data demands the introduction of algorithmic
approaches.

The development of automatic identification and quantification of cell
type may also provide a new computational phenotype for quantifying
the effect of disease, genetic interventions, and
developmentally-experienced neural activity. Our method can in
principle identify neuron-types across non-connected graphs,
e.g. across animals. For example, the types of neurons in one animal
can be associated with the types of neurons in another animal, in the
same way as this is already possible through molecular marker
\autocite{Brown2009}. This could be particularly important if cell
types appear that are due to properties of the stimuli and experience
as opposed to just the molecular properties of cells, such as color
and orientation selective types in primary visual cortex
\autocite{Sincich2005,Lennie2005}. This would allow comparative
quantitative anatomy across animals, and aid the search for the
ultimate causes of connectivity.

Our model combines connectivity, cellular, and synaptic properties, 
and suggests the way towards combining even richer data. Distinct cell
types differ in morphology, connectivity, transcriptomics, relation to
behavior or stimuli and many other ways. Algorithms combining this data and type type
information may allow us to synthesize all the available
information from one experiment or even across experiments into a
joint model of brain structure and function.

Eventually, algorithms will have to become a central tool for
anatomists, as it will progressively become impossible for humans to
parse the huge datasets. This transition may follow a similar
transition to the one underway in molecular biology (with gene-finding
algorithms) and evolutionary biology with (computational
phylogenetics).  In fact, it is known that there are significant
disagreements across human anatomists and computational approaches
promise to make the debates more data driven.

@@

For the basic link-distance model, we take as input a connectivity
matrix $R$ defining the connections between cell $e_i$ and $e_j$, as
well as a distance function $d(e_i, e_j)$ representing a (physical)
distance between adjacent cells. \todo{Add the cell soma caveat?}  See
the supplemental material for extension to multiple connectivity
matrices. We assume there exist an unknown number $K$ of latent
(unobserved) cell types, $k \in \{1, 2, 3, \dots, K\}$, and that each
cell $e_i$ belongs to a single cell type. We indicate a cell $e_i$ is
of type $k$ using the assignment vector $\vec(c)$, so $c_i = k$. The
observed connectivity between two cells $R(e_i, e_j)$ then depends
only on their latent type and their distance through a link function
$f(\cdot, d(e_i, e_j))$. We assume $f$ is parameterized based on the
latent type, $c_i=m$ and $c_j=n$, via a parameter$\eta_{mn}$, as well
as a set of global hyper parameters $\theta$, such that the link
function is $f(d(e_i, e_j) | \eta_{mn}, \theta)$.

@@

For the retina data, we then extend the model with the additional
features indicated. Cell soma depth is modeled as a
cell-type-dependent Gaussian with latent (unknown) per-type mean and
variance. Similarly, each cell has some number $N_i$ of synapses, 
each of which is drawn from a cell-type-specific density profile
with up to three modes.

@@

\newpage
\section{Methods }
XXX TODO : add n-way cross validation is not done because it would be a pain
XXX TODO that said overfitting may be shared across all models

@@

% TODO add in the thingie
\begin{equation}
  p(\vec{c}, \eta, \theta | R ) \propto \prod_{i, j} p(R(e_i, e_j) | f(d(e_i, e_j) | \eta_{c_ic_j}), \theta) \prod_{m, n} p(\eta_{mn} | \theta)  p(\theta) p(\vec{c} | \alpha) p(\alpha) p(\theta)
\end{equation}

@@

Where we model cell depth, we assume that 
each cell type has a typical depth, and thus a Gaussian distribution
of $s_i$. We assume $s_i \sim N(\mu^{(s)}_k,
\sigma^{2(s)}_k)$, where the $(s)$ superscript indicates
these model parameters are associated with the soma-depth 
portion of our model.  We use a conjugate prior for $(\mu^{(s)}_k, \sigma^{2(s)}_k)$ with
$\mu^{(s)}_k \sim N(\mu^{(s)}_{hp}, \sigma^{2(s)}_k/ \kappa^{(s)}_{hp})$ and $\sigma^{2(s)}_k \sim
\chi^{-1}(\sigma^{2(s)}_{hp}, \nu^{(s)}_{hp}$. The use of conjugacy simplifies inference while allowing for each cell-type to have its own depth mean and distribution.

Where we model synapse depth profile, we assume that each cell
type has a characteristic depth distribution of synaptic contact points, and thus
a mixture of Gaussians distribution over cell $i$’s $N_i$ contact points, $\vec{g^i}$.
We do this by assuming the $g^i_j$
are drawn from an $M=3$-component mixture of Gaussians. Thus associated with each cell type $k$ is a vector
of $M$ Gaussian means $(\mu^g_{k,1}, \cdots, \mu^g_{k, M})$, and a
mixture vector $\pi_k$.  This representation can thus model depth distributions of contact points that have up to three modes, an assumption that is well matched in the bulk of anatomical studies of cell-type dependent connectivity.

@@

To pick the MAP, we run up to $100$ simultaneous chains in parallel,
initialized from different random initial points in the state space. 
For visualization we pick the chain with the highest log likelihood,
but for all numerical comparisons (including link probability and ARI)
we use this full collection of posterior samples to estimate
the resulting statistics.

\subsubsection{Mixing of our Markov chains} 
Evaluating whether or not approximate inference methods,
such as MCMC, produce samples which are valid approximations 
of the posterior distribution is an ongoing area of research
in the computational statistics community. Here we validate
mixing in two ways:

\being{itemize}
\item \textbf{synthetic}: For synthetic datasets of sizes comparable to our
real data size, do we recover known ground truth information
after running our markov chains for the appropriate amount of time? 
\todo{Add figure}
\item \textbf{autocorrelation} For various summary statistics 
of the latent space, after a certain amount of time 
\todo{add figure}
\end{itemize}

But importantly, regardless of whether our model over- or under-estimates the exact posterior variance about the network, we find points in the latent variable space that are both predictive \textit{and} parsimonious, largely agreeing with the human anatomists and predicting existing connections.

@@

% \begin{addendum}
%  \item Put acknowledgements here.
%  \item[Competing Interests] The authors declare that they have no
% competing financial interests.
%  \item[Correspondence] Correspondence and requests for materials
% should be addressed to A.B.C.~(email: myaddress@nowhere.edu).
% \end{addendum}
\newpage
\section{Supplemental Material}
